\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-07 2-the-cubble-pkg.Rmd
some changes as per meeting today

Spatio-temporal data usually come to the analysts in various forms and Figure \ref{fig:illu-input} shows four examples of this. These four examples show the common characters of the spatio-temporal data: 1) a spatial identifier that uniquely identifies each site, 2) a temporal identifier that prescribes the time stamp each site is recorded, 3) Coordinates, comprising of latitude and longitude, that locate the site on the map. These identifiers will be the building blocks for cubble. Another way to structure the spatio-temporal data is to think of it as a data cube with three axes: id, time, and variable. Notice that in the time-variable face of this cube, time invariant variables would share the same value across all the time stamps, while time varying variables have values varying across time. (Maybe add a cube diagram here). This heterogeneity will be used to split the data into two forms in cubble.

@@

For those operations whose output involves both a spatial and temporal identifier, long form should be used. The long form uses a  \code{`grouped_df`} data frame to forms all the time of a site as a group. Time invariant variables are stored separately as an special attribute of the long cubble. This design avoids repeating the spatial variables at each time stamp while not dropping information from spatial variables.

@@

There are a few information in the cubble header: the name of the \code{key} variable (`id`) and its number (5), the bounding box: \code{[115.97, -32.94, 133.55, -12.42]} , and also the name of variable nested in the \code{ts} column with its type: \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the data and  \code{stretch()} switches the nested cubble into the long cubble. It first extracts all the spatial variables into a separate tibble and then unnests the \code{ts} column. The spatial variables are then attached as an attributes of the long cubble and can be accessed through `spatial()`

@@

Analysts may need the time invariant/ varying variable to be in the same table to apply certain function. In cubble, this can be done through \code{migrate()}, which moves the time invariant variables into the long cubble:

@@

Notice that these added time invariant variables will not be remembered by cubble and will disappear if switched to the nested form and then switched back. Hence, \code{migrat} should usually be used as the last step of the analysis:

@@

@@

\pkg{cubble} is also compatible with \pkg{tsibble}. When creating a cubble from a tsibble, Only the \code{coords} argument need to be specified:

@@

>>>>

Spatio-temporal data usually come to the analysts in various forms and Figure \ref{fig:illu-input} shows four examples of this. The first three are variations of tabular data that uses a single table (1) or multiple tables (2 & 3) while the fourth uses a cube or hyper-cube to to represent the data. An example of the fourth type is the NetCDF (Network Common Data Form), commonly used in the climatology community to deliver global mapping of atmosphere, ocean, and land.

These four examples show the common characters of the spatio-temporal data: 1) a spatial identifier that uniquely identifies each site, 2) a temporal identifier that prescribes the time stamp each site is recorded, 3) Coordinates, comprising of latitude and longitude, that locate the site on the map. These identifiers will be the building blocks for cubble. Another way to structure the spatio-temporal data is to think of it as a data cube with three axes: id, time, and variable. Notice that in the time-variable face of this cube, time invariant variables would share the same value across all the time stamps, while time varying variables have values varying across time. (Maybe add a cube diagram here). This heterogeneity will be used to split the data into two forms in cubble.

@@

For those operations whose output involves both a spatial and temporal identifier, long form should be used. The long form uses a  \code{grouped_df} data frame to forms all the time of a site as a group. Time invariant variables are stored separately as an special attribute of the long cubble. This design avoids repeating the spatial variables at each time stamp while not dropping information from spatial variables.

@@

There are a few information in the cubble header: the name of the \code{key} variable: \code{id} and its number: 5, the bounding box: \code{[115.97, -32.94, 133.55, -12.42]} , and also the name of variable nested in the \code{ts} column with its type: \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the data and  \code{stretch()} switches the nested cubble into the long cubble. It first extracts all the spatial variables into a separate tibble and then unnests the \code{ts} column. The spatial variables are then attached as an attributes of the long cubble and can be accessed through `spatial()`:

@@

Analysts may need the time invariant and varying variable to be in the same table to apply certain function. In cubble, this can be done through \code{migrate()}, which moves the time invariant variables into the long cubble:

@@

Notice that these added time invariant variables will not be remembered by cubble and will disappear if switched to the nested form and then switched back. Hence, \code{migrate()} should usually be used as the last step of the analysis:

@@

### dplyr

@@

### tsibble

Cubble is also compatible with \pkg{tsibble}. When creating a cubble from a tsibble, Only the \code{coords} argument needs to be specified:

@@

### sf

\textcolor{red}{add content}

### netcdf

NetCDF data has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes are usually associated with dimension and variable in the NetCDF format data and a [metadata convention for climate and forecast](http://cfconventions.org/) has been designed to standardise the format of the attributes. A few packages in R exists for manipulating NetCDF data and this includes a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls C-interface: \pkg{RNetCDF} [@rnetcdf; @michna2013rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

Cubble provides an \code{as_cubble()} method to coerce the \code{ncdf4} class from the \pkg{ncdf4} package into a \code{cubble}. It maps each combination of longitude and latitude into an \code{id} as the \code{key}:

```{r echo = TRUE, eval = FALSE}
# read in the .nc file as a ncdf4 class
raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))

# convert the variable q and z in the ncdf4 into a cubble
dt <- as_cubble(raw, vars = c("q", "z"))
```

Memory limit with NetCDF data in cubble depends on longitude grid point x latitude grid point x time grid point x number of variable. Cubble can handle slightly more than 300 x 300 (longitude x longitude) grid points for 3 variables in one year and spatial grid can be reduced to trade for longer time period and more variables. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or a global bounding box [-180, -90, 180, -90] at 1 degree resolution. Subsetting longitude and latitude grid is available through \code{long_range} and \code{lat_range} if the NetCDF file has finer resolution than needed.

```{r eval = FALSE, echo = TRUE}
# Assume my_ncdf has bounding box of [-180, -90, 180, -90] 
# at 0.25 degree resolution and subset it to have 
# 1 degree resolution:
dt <- as_cubble(my_ncdf, vars = c("q", "z"), 
                long_range = seq(-180, 180, 1), 
                lat_range = seq(-90, 90, 1))
```

### 2022-02-07 4-3-water-level.Rmd
some changes as per meeting today

## River level data in Victria water gauges

@@

From the map, a few water gauges and weather stations are close to each other and the fluctuation of the water level could be matched up with precipitation measured by the climate station. As introduced in Section 3.2, \code{match_sites()} can be used to match one source of data with another source in a cubble. Here \code{Water_course_level} in \code{river} will be matched to \code{prcp} in \code{climate} in 2020. The two datasets need to be specified as the first two arguments and the variable to match can be specified in \code{temporal_var_to_match} using the \code{by} syntax in \code{join}. \code{temporal_independent} controls the variable used to construct the interval and the goal here is to see if precipitation will be reflected by the water level in the river. This puts precipitation \code{prcp}, as the independent. Given there is one year worth of data, the number of peak (\code{temporal_n_highest}) to consider is slightly raised from a default 20 to 30 and \code{temporal_min_match} is raised correspondingly. To return all the pairs of the match, \code{temporal_min_match} can be set to 0.

```{r echo = TRUE}
res <- match_sites(
  river, climate,
  temporal_var_to_match = c("Water_course_level" = "prcp"),
  temporal_independent = "prcp",  
  temporal_n_highest = 30,
  temporal_min_match = 15
)
```

>>>>

## River level data in Victoria water gauges

@@

From the map, a few water gauges and weather stations are close to each other and the fluctuation of the water level could be matched up with precipitation measured by the climate station. As introduced in Section 3.2, \code{match_sites()} can be used to match one source of data with another source in a cubble. Here \code{Water_course_level} in \code{river} will be matched to \code{prcp} in \code{climate} in 2020. The two datasets need to be specified as the first two arguments and the variable to match can be specified in \code{temporal_by} using the \code{by} syntax in \code{join}. \code{temporal_independent} controls the variable used to construct the interval and the goal here is to see if precipitation will be reflected by the water level in the river. This puts precipitation \code{prcp}, as the independent. Given there is one year worth of data, the number of peak (\code{temporal_n_highest}) to consider is slightly raised from a default 20 to 30 and \code{temporal_min_match} is raised correspondingly. To return all the pairs of the match, \code{temporal_min_match} can be set to 0.

```{r echo = TRUE}
res <- match_sites(
  river, climate,
  temporal_by = c("Water_course_level" = "prcp"),
  temporal_independent = "prcp",  
  temporal_n_highest = 30,
  temporal_min_match = 15
)
```

### 2022-02-07 4-4-netcdf.Rmd
some changes as per meeting today

NetCDF (Network Common Data Form) is a data format commonly used in the climatology community to deliver global mapping of atmosphere, ocean, and land. It has two major components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes are usually associated with dimension and variable in the NetCDF format data and a [metadata convention for climate and forecast](http://cfconventions.org/) has been designed to standardise the format of the attributes. Current packages for manipulating NetCDF data in R include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls C-interface: \pkg{RNetCDF} [@rnetcdf; @michna2013rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

Cubble provides an \code{as_cubble()} method to coerce the \code{ncdf4} class from the \pkg{ncdf4} package into a \code{cubble}. It maps each combination of longitude and latitude into an \code{id} as the \code{key}:

```{r echo = TRUE, eval = FALSE}
# read in the .nc file as a ncdf4 class
raw <- ncdf4::nc_open(here::here("data/era5-pressure.nc"))

# convert the variable q and z in the ncdf4 into a cubble
dt <- as_cubble(raw, vars = c("q", "z"))
```

Memory limit with NetCDF data in cubble depends on longitude grid point x latitude grid point x time grid point x number of variable. Cubble can handle slightly more than 300 x 300 (longitude x longitude) grid points for 3 variables in one year and spatial grid can be reduced to trade for longer time period and more variables. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or a global bounding box [-180, -90, 180, -90] at 1 degree resolution. Subsetting longitude and latitude grid is available through \code{long_range} and \code{lat_range} if the NetCDF file has finer resolution than needed.

```{r eval = FALSE, echo = TRUE}
# Assume my_ncdf has bounding box of [-180, -90, 180, -90] 
# at 0.25 degree resolution and subset it to have 
# 1 degree resolution:
dt <- as_cubble(my_ncdf, vars = c("q", "z"), 
                long_range = seq(-180, 180, 1), 
                lat_range = seq(-90, 90, 1))
```

Figure \ref{fig:netcdf} gives an example of reproducing Figure 19 in @hersbach2020era5 using ERA5 data. ERA5 data [@hersbach2020era5] is the latest reanalysis of global atmosphere, land surface, and ocean waves from 1950 onwards and is available in the NetCDF format from European Centre for Medium-Range Weather Forecasts (ECMWF). The data can be directly downloaded from [Copernicus Climate Data Store (CDS)](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels?tab=overview) website or programmatically via an R package \pkg{ecmwfr} [@ecwmfr]. Variable *specific humidity* and *geopotential* are queried on the 10 hPa pressure level on four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once downloaded, the data can be read into a cubble as:

@@

Readers who are interested in the analysis of the figure can refer to @simmons2020global and @simmons2005ecmwf for more details.

>>>>

ERA5 data [@hersbach2020era5] is the latest reanalysis of global atmosphere, land surface, and ocean waves from 1950 onwards and is available in the NetCDF format from European Centre for Medium-Range Weather Forecasts (ECMWF). The data can be directly downloaded from [Copernicus Climate Data Store (CDS)](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels?tab=overview) website or programmatically via an R package \pkg{ecmwfr} [@ecwmfr]. The \code{era5-pressure} data contains variable *specific humidity* and *geopotential* on the 10 hPa pressure level on four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once downloaded, the data can be read into a cubble as:

@@

Figure \ref{fig:netcdf} reproduces the ERA5 data row of Figure 19 in @hersbach2020era5. It shows the southern polar vortex splits into two on 2002-09-26 and further splits into four on 2002-10-04 in the stratosphere. Readers interested in the analysis of this figure can refer to @hersbach2020era5, @simmons2020global and @simmons2005ecmwf for more details.

### 2022-02-07 main.Rmd
some changes as per meeting today

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal variables can be divided into time invariant ones that can be identified by the spatial identifier, or the time varying ones that cross-identified by the spatial and temporal identifiers. In this paper, a new data structure, \code{cubble}, is proposed for manipulating and visualising spatio-temporal data. The new data structure uses two forms to organise time invariant/ varying variables so that the spatial and temporal dimension of the data can be manipulated separately while keep synchronised. Advanced considerations are given to illustrate data with hierarchical structure, data matching, and how cubble fits with the interactive graphics pipeline. Examples are given to analysing Australia meteorology data, river level data, and climate reanalysis (ERA5). 
keywords:
  # at least one keyword must be supplied
  formatted: [spatio temporal,  "\\proglang{R}", climate weather station, data analysis]
  plain:     [spatio temporal, R, climate weather station, data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal variables can be divided into time invariant ones that can be identified by the spatial identifier, or the time varying ones that cross-identified by the spatial and temporal identifiers. In this paper, a new data structure, \code{cubble}, is proposed for manipulating and visualising spatio-temporal data. The new data structure uses two forms to organise time invariant and varying variables so that the spatial and temporal dimension of the data can be manipulated separately while keep synchronised. Advanced considerations are given to illustrate data with hierarchical structure, data matching, and how cubble fits with the interactive graphics pipeline. Examples are given to analysing Australia meteorology data, river level data, and climate reanalysis (ERA5). 
keywords:
  # at least one keyword must be supplied
  formatted: [spatio temporal,  "\\proglang{R}", climate weather station, data analysis]
  plain:     [spatio temporal, R, climate weather station, data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

### 2022-02-07 main.tex
some changes as per meeting today

\Abstract{
Spatio-temporal variables can be divided into time invariant ones that
can be identified by the spatial identifier, or the time varying ones
that cross-identified by the spatial and temporal identifiers. In this
paper, a new data structure, \code{cubble}, is proposed for manipulating
and visualising spatio-temporal data. The new data structure uses two
forms to organise time invariant/ varying variables so that the spatial
and temporal dimension of the data can be manipulated separately while
keep synchronised. Advanced considerations are given to illustrate data
with hierarchical structure, data matching, and how cubble fits with the
interactive graphics pipeline. Examples are given to analysing Australia
meteorology data, river level data, and climate reanalysis (ERA5).
}

@@

Spatio-temporal data usually come to the analysts in various forms and
Figure \ref{fig:illu-input} shows four examples of this. These four
examples show the common characters of the spatio-temporal data: 1) a
spatial identifier that uniquely identifies each site, 2) a temporal
identifier that prescribes the time stamp each site is recorded, 3)
Coordinates, comprising of latitude and longitude, that locate the site
on the map. These identifiers will be the building blocks for cubble.
Another way to structure the spatio-temporal data is to think of it as a
data cube with three axes: id, time, and variable. Notice that in the
time-variable face of this cube, time invariant variables would share
the same value across all the time stamps, while time varying variables
have values varying across time. (Maybe add a cube diagram here). This
heterogeneity will be used to split the data into two forms in cubble.

@@

There are a few information in the cubble header: the name of the
\code{key} variable (\texttt{id}) and its number (5), the bounding box:
\code{[115.97, -32.94, 133.55, -12.42]} , and also the name of variable
nested in the \code{ts} column with its type:
\code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the data
and \code{stretch()} switches the nested cubble into the long cubble. It
first extracts all the spatial variables into a separate tibble and then
unnests the \code{ts} column. The spatial variables are then attached as
an attributes of the long cubble and can be accessed through
\texttt{spatial()}

@@

@@

\pkg{cubble} is also compatible with \pkg{tsibble}. When creating a
cubble from a tsibble, Only the \code{coords} argument need to be
specified:

@@

\hypertarget{advanced-features-considerations}{%
\section{Advanced features/
considerations}\label{advanced-features-considerations}}

@@

\hypertarget{glyph-map}{%
\subsection{Glyph map}\label{glyph-map}}

@@

\hypertarget{river-level-data-in-victria-water-gauges}{%
\subsection{River level data in Victria water
gauges}\label{river-level-data-in-victria-water-gauges}}

@@

From the map, a few water gauges and weather stations are close to each
other and the fluctuation of the water level could be matched up with
precipitation measured by the climate station. As introduced in Section
3.2, \code{match_sites()} can be used to match one source of data with
another source in a cubble. Here \code{Water_course_level} in
\code{river} will be matched to \code{prcp} in \code{climate} in 2020.
The two datasets need to be specified as the first two arguments and the
variable to match can be specified in \code{temporal_var_to_match} using
the \code{by} syntax in \code{join}. \code{temporal_independent}
controls the variable used to construct the interval and the goal here
is to see if precipitation will be reflected by the water level in the
river. This puts precipitation \code{prcp}, as the independent. Given
there is one year worth of data, the number of peak
(\code{temporal_n_highest}) to consider is slightly raised from a
default 20 to 30 and \code{temporal_min_match} is raised
correspondingly. To return all the pairs of the match,
\code{temporal_min_match} can be set to 0.

\begin{CodeChunk}
\begin{CodeInput}
R> res <- match_sites(
+   river, climate,
+   temporal_var_to_match = c("Water_course_level" = "prcp"),
+   temporal_independent = "prcp",  
+   temporal_n_highest = 30,
+   temporal_min_match = 15
+ )
\end{CodeInput}
\end{CodeChunk}

@@

NetCDF (Network Common Data Form) is a data format commonly used in the
climatology community to deliver global mapping of atmosphere, ocean,
and land. It has two major components: \emph{dimension} for defining the
spatio-temporal grid (longitude, latitude, and time) and \emph{variable}
that populates the defined grid. Attributes are usually associated with
dimension and variable in the NetCDF format data and a
\href{http://cfconventions.org/}{metadata convention for climate and
forecast} has been designed to standardise the format of the attributes.
Current packages for manipulating NetCDF data in R include a high-level
R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls
C-interface: \pkg{RNetCDF} \citep{rnetcdf, michna2013rnetcdf}, and a
tidyverse implementation: \pkg{tidync} \citep{tidync}.

Figure \ref{fig:netcdf} gives an example of reproducing Figure 19 in
\citet{hersbach2020era5} using ERA5 data. ERA5 data
\citep{hersbach2020era5} is the latest reanalysis of global atmosphere,
land surface, and ocean waves from 1950 onwards and is available in the
NetCDF format from European Centre for Medium-Range Weather Forecasts
(ECMWF). The data can be directly downloaded from
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus
Climate Data Store (CDS)} website or programmatically via an R package
\pkg{ecmwfr} \citep{ecwmfr}. Variable \emph{specific humidity} and
\emph{geopotential} are queried on the 10 hPa pressure level on four
dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once
downloaded, the data can be read into a cubble as:

@@

Readers who are interested in the analysis of the figure can refer to
\citet{simmons2020global} and \citet{simmons2005ecmwf} for more details.

@@

This paper describes an \proglang{R} package \pkg{cubble} for
manipulating and visualising spatio-temporal data. A new data structure,
\code{cubble} that builds from the \code{rowwise_df} and
\code{grouped_df} class in the tidyverse ecosystem, is proposed to
connect the time invariant and varying variables in the spatio-temporal
data. This design frees the data analysts from spending time on
organising variables of different observational units. The data
structure is also flexible to the techniques and packages analysts use
to analyse the data, for example, in the matching example in section
4.3, users are free to use algorithms from another package to cluster
stations. Further development and maintenance of the package involves
responding to changes in the tidyverse packages that \pkg{cubble}
imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}.
Another area for further development is to extend \pkg{cubble} to other
spatial objects other than points.

>>>>

\Abstract{
Spatio-temporal variables can be divided into time invariant ones that
can be identified by the spatial identifier, or the time varying ones
that cross-identified by the spatial and temporal identifiers. In this
paper, a new data structure, \code{cubble}, is proposed for manipulating
and visualising spatio-temporal data. The new data structure uses two
forms to organise time invariant and varying variables so that the
spatial and temporal dimension of the data can be manipulated separately
while keep synchronised. Advanced considerations are given to illustrate
data with hierarchical structure, data matching, and how cubble fits
with the interactive graphics pipeline. Examples are given to analysing
Australia meteorology data, river level data, and climate reanalysis
(ERA5).
}

@@

Spatio-temporal data usually come to the analysts in various forms and
Figure \ref{fig:illu-input} shows four examples of this. The first three
are variations of tabular data that uses a single table (1) or multiple
tables (2 \& 3) while the fourth uses a cube or hyper-cube to to
represent the data. An example of the fourth type is the NetCDF (Network
Common Data Form), commonly used in the climatology community to deliver
global mapping of atmosphere, ocean, and land.

These four examples show the common characters of the spatio-temporal
data: 1) a spatial identifier that uniquely identifies each site, 2) a
temporal identifier that prescribes the time stamp each site is
recorded, 3) Coordinates, comprising of latitude and longitude, that
locate the site on the map. These identifiers will be the building
blocks for cubble. Another way to structure the spatio-temporal data is
to think of it as a data cube with three axes: id, time, and variable.
Notice that in the time-variable face of this cube, time invariant
variables would share the same value across all the time stamps, while
time varying variables have values varying across time. (Maybe add a
cube diagram here). This heterogeneity will be used to split the data
into two forms in cubble.

@@

There are a few information in the cubble header: the name of the
\code{key} variable: \code{id} and its number: 5, the bounding box:
\code{[115.97, -32.94, 133.55, -12.42]} , and also the name of variable
nested in the \code{ts} column with its type:
\code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the data
and \code{stretch()} switches the nested cubble into the long cubble. It
first extracts all the spatial variables into a separate tibble and then
unnests the \code{ts} column. The spatial variables are then attached as
an attributes of the long cubble and can be accessed through
\texttt{spatial()}:

@@

\hypertarget{dplyr}{%
\subsubsection{dplyr}\label{dplyr}}

@@

\hypertarget{tsibble}{%
\subsubsection{tsibble}\label{tsibble}}

Cubble is also compatible with \pkg{tsibble}. When creating a cubble
from a tsibble, Only the \code{coords} argument needs to be specified:

@@

\hypertarget{sf}{%
\subsubsection{sf}\label{sf}}

\textcolor{red}{add content}

\hypertarget{netcdf}{%
\subsubsection{netcdf}\label{netcdf}}

NetCDF data has two main components: \emph{dimension} for defining the
spatio-temporal grid (longitude, latitude, and time) and \emph{variable}
that populates the defined grid. Attributes are usually associated with
dimension and variable in the NetCDF format data and a
\href{http://cfconventions.org/}{metadata convention for climate and
forecast} has been designed to standardise the format of the attributes.
A few packages in R exists for manipulating NetCDF data and this
includes a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a
low-level interface that calls C-interface: \pkg{RNetCDF}
\citep{rnetcdf, michna2013rnetcdf}, and a tidyverse implementation:
\pkg{tidync} \citep{tidync}.

\hypertarget{other-features-and-considerations}{%
\section{Other features and
considerations}\label{other-features-and-considerations}}

@@

\hypertarget{spatio-temporal-transformations}{%
\subsection{Spatio-temporal
transformations}\label{spatio-temporal-transformations}}

\textcolor{red}{Glyph map + transforming on time (calendar plot) + projections (linear combination).}

@@

\hypertarget{river-level-data-in-victoria-water-gauges}{%
\subsection{River level data in Victoria water
gauges}\label{river-level-data-in-victoria-water-gauges}}

@@

From the map, a few water gauges and weather stations are close to each
other and the fluctuation of the water level could be matched up with
precipitation measured by the climate station. As introduced in Section
3.2, \code{match_sites()} can be used to match one source of data with
another source in a cubble. Here \code{Water_course_level} in
\code{river} will be matched to \code{prcp} in \code{climate} in 2020.
The two datasets need to be specified as the first two arguments and the
variable to match can be specified in \code{temporal_by} using the
\code{by} syntax in \code{join}. \code{temporal_independent} controls
the variable used to construct the interval and the goal here is to see
if precipitation will be reflected by the water level in the river. This
puts precipitation \code{prcp}, as the independent. Given there is one
year worth of data, the number of peak (\code{temporal_n_highest}) to
consider is slightly raised from a default 20 to 30 and
\code{temporal_min_match} is raised correspondingly. To return all the
pairs of the match, \code{temporal_min_match} can be set to 0.

\begin{CodeChunk}
\begin{CodeInput}
R> res <- match_sites(
+   river, climate,
+   temporal_by = c("Water_course_level" = "prcp"),
+   temporal_independent = "prcp",  
+   temporal_n_highest = 30,
+   temporal_min_match = 15
+ )
\end{CodeInput}
\end{CodeChunk}

@@

ERA5 data \citep{hersbach2020era5} is the latest reanalysis of global
atmosphere, land surface, and ocean waves from 1950 onwards and is
available in the NetCDF format from European Centre for Medium-Range
Weather Forecasts (ECMWF). The data can be directly downloaded from
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus
Climate Data Store (CDS)} website or programmatically via an R package
\pkg{ecmwfr} \citep{ecwmfr}. The \code{era5-pressure} data contains
variable \emph{specific humidity} and \emph{geopotential} on the 10 hPa
pressure level on four dates: 2002-09-22, 2002-09-26, 2002-09-30, and
2002-10-04. Once downloaded, the data can be read into a cubble as:

@@

Figure \ref{fig:netcdf} reproduces the ERA5 data row of Figure 19 in
\citet{hersbach2020era5}. It shows the southern polar vortex splits into
two on 2002-09-26 and further splits into four on 2002-10-04 in the
stratosphere. Readers interested in the analysis of this figure can
refer to \citet{hersbach2020era5}, \citet{simmons2020global} and
\citet{simmons2005ecmwf} for more details.

@@

This paper describes an \proglang{R} package \pkg{cubble} for
manipulating and visualising spatio-temporal data. A new data structure,
\code{cubble} that builds from the \code{rowwise_df} and
\code{grouped_df} class in the tidyverse ecosystem, is proposed to
connect the time invariant and varying variables in the spatio-temporal
data. This design frees the data analysts from spending time on
organising variables of different observational units. The data
structure is also flexible to the techniques and packages analysts use
to analyse the data, for example, in the matching example in section
4.3, users are free to use algorithms from another package to cluster
stations.

Further development and maintenance of the package involves responding
to changes in the tidyverse packages that \pkg{cubble} imports, in
particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. Another area for
further development is to extend \pkg{cubble} to other spatial objects
other than points.
